Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-03-07 manuscript/manuscript.Rmd
Pre-Anitra edits v2

High frequency, long-term observations at Station ALOHA in the North Pacific Subtropical Gyre (NPSG) over the past 25 years have revealed temporal and spatial variability in what had prevously been considered a homogenous environment [@Karl1999; @Karl2017; @Karl2021]. A major source of this variability comes in the form of mesoscale eddies in which water is entrained into circular surface currents tens to hundreds of kilometers in diameter [@Karl2017; @McGillicuddy2016]. Prior research has established the importance of these mesoscale features in introducing nutrients into the euphotic zone and altering the community composition [@Sweeney2003; @Barone2019]. In particular, the depth and intensity of the deep chlorophyll maximum (DCM) at the top of the nitricline differs dramatically between cyclonic and anticyclonic eddies. @Barone2022 reported a shift in the microbial community at the DCM with eukaryotic phytoplankton enriched and prokaryotic phytoplankton depleted in cyclonic eddies, and @Hawco2021 and @Dugenne2023 showed that some of this difference was due to alterations in dissolved iron chemistry and nitrogen fixation.

@@

```{r figure stacked barplot MT eddy transect w PC and SLA}
avg_targ_concs <- targ_peaks %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MT") %>%
  filter(samp_type=="Smp") %>%
  group_by(compound_name) %>%
  mutate(mean_cmpd_nm=mean(nM)) %>%
  ungroup()
arrow_df <- data.frame(
  cyc_type=c("Anticyclone\ncenter", "Cyclone\ncenter"),
  x=c("6", "12"), xend=c("6", "12"),
  y=c(5, 4), yend=c(9.5, 6.5),
  color=c(scales::muted("red"), scales::muted("blue")),
  depth=factor("Deep", levels=c("Surface", "DCM", "Deep"), 
               labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)")),
  lab_x=c("5", "10")
)
barplot_targ_gp <- avg_targ_concs %>%
  group_by(compound_name, untripl) %>%
  summarise(nM=median(nM), mean_cmpd_nm=unique(mean_cmpd_nm)) %>%
  ungroup() %>%
  arrange(desc(mean_cmpd_nm)) %>%
  mutate(compound_name=ifelse(compound_name=="Trimethylamine N-oxide", "TMAO", compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Hydroxyisoleucine", "HO-Ile", compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Glycine betaine", "GBT", compound_name)) %>%
  mutate(compound_name=str_remove(compound_name, "^L-")) %>%
  mutate(compound_name=ifelse(
    compound_name%in%head(unique(compound_name), 9), 
    compound_name, "Other")) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  left_join(filled_file_metadata %>% distinct(untripl, depth, station)) %>%
  group_by(depth, station, compound_name) %>%
  summarise(nM=sum(nM), .groups = "keep") %>%
  summarise(mean_nM=mean(nM)) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)"))) %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_col(aes(x=station, y=mean_nM, fill=compound_name), width = 0.9) +
  scale_fill_manual(values = c(splividis(9), "grey50")) +
  scale_y_continuous(expand = c(0.005, 0.1)) +
  labs(x=NULL, y="Median nM\nmetabolite concentration", fill=NULL) +
  facet_grid(~depth, scales="free", space = "free") +
  guides(fill=guide_legend(nrow=2, byrow = TRUE)) +
  ggnewscale::new_scale_fill() +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend, color=color), data=arrow_df,
               arrow=arrow(ends = "first", length = unit(0.1, "npc"), type = "closed"),
               linewidth=0, linejoin = "bevel") +
  geom_segment(aes(x=x, y=y+0.5, xend=xend, yend=yend, color=color), 
               data=arrow_df, linewidth=2) +
  geom_text(aes(x=lab_x, y=yend+0.5, color=color, label=cyc_type), 
            data=arrow_df, vjust=0, hjust=0, lineheight=0.8) +
  scale_x_discrete(breaks=as.character(seq(4, 14, 2))) +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("fill"), guide = "none"
  ) +
  scale_color_identity() +
  theme_bw() +
  theme(legend.position = "top", axis.text.x = element_blank(), 
        legend.margin = margin(t = 10,0,0,0), 
        legend.box.margin = margin(-10,-10,-10,-10),
        plot.margin = unit(numeric(4), "cm"),
        strip.text = element_text(face = "bold"))

frac_pc_in_metab_df <- targ_peaks %>%
  filter(str_detect(filename, "180821_Smp")) %>%
  left_join(clean_stans %>% select(compound_name, formula, polarity)) %>%
  mutate(n_C=as.numeric(str_extract(formula, "(?<=C)\\d+"))) %>%
  mutate(nM_C=n_C*nM) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  filter(samp_type=="Smp") %>%
  group_by(filename) %>%
  summarize(total_nM=sum(nM_C), PC_um=unique(PC_um)) %>%
  mutate(frac_in_metab=total_nM/1000/PC_um) %>%
  left_join(filled_file_metadata, by = join_by(filename, PC_um)) %>%
  mutate(depth=str_extract(depth, "DCM|Deep|Surface")) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  mutate(station=factor(station, levels=c(as.character(4:14))))
pc_frac_metab_gp <- frac_pc_in_metab_df %>%
  ggplot(aes(x=station, y=frac_in_metab, color=sla_corr)) +
  geom_hline(yintercept = 0) +
  geom_boxplot(width = 0.9, linewidth=0.8) +
  scale_x_discrete(breaks=as.character(seq(4, 14, 2))) +
  scale_y_continuous(labels = scales::label_percent()) +
  facet_grid(~depth, scales="free") +
  scale_color_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("color", "fill")
  ) +
  guides(color=guide_colorbar(direction = "horizontal", title.position = "top", 
                              title.hjust = 0.5, barheight = 0.2, reverse = TRUE)) +
  labs(color="Corr. SLA (cm)", x="Station",
       y="Fraction of PC represented\nby known metabolites") +
  theme_bw() +
  theme(legend.position = c(0.84, 0.75), legend.background = element_rect(color="grey90"),
        strip.background = element_blank(), strip.text = element_blank(),
        plot.margin = margin(0,1,0,0))

@@

no_outlier_frac_pc_in_metab_df <- frac_pc_in_metab_df %>%
  filter(!(station=="11" & tripl=="A" & depth=="Surface")) %>%
  filter(!(station=="6" & tripl=="B" & depth=="Surface")) %>%
  filter(!(station=="7" & tripl=="C" & depth=="Surface")) %>%
  filter(!(station=="11" & tripl=="B" & depth=="DCM"))
pc_totalnm_plot <- no_outlier_frac_pc_in_metab_df %>%
  ggplot() +
  aes(x=PC_um, y=total_nM) +
  geom_smooth(method="lm", color="black") +
  geom_point(aes(fill=sla_corr), pch=21, color="black") +
  facet_wrap(~depth, ncol=1, scales="free") +
  labs(x="Particulate\ncarbon (uM)", y="Total nM carbon in known metabolites") +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("fill"), guide = "none"
  ) +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))

targ_gp_w_sla_pc <- cowplot::plot_grid(
  targ_gp_w_sla_frac, pc_totalnm_plot, rel_widths = c(3, 1), labels = "AUTO"
) +
  theme(plot.background = element_rect(fill="white", color=NA))

plot(targ_gp_w_sla_pc)

ggsave("targ_gp_w_sla_pc.png", plot = targ_gp_w_sla_pc, device = "png", 
       path = "figures", width = 6.875, height = 4.5, units = "in", dpi = 300)
```

![](figures/targ_gp_w_sla_pc.png)

*Figure 2: A) Upper row of stacked barplots shows known compounds measured across the eddy dipole and separated by depth. Bar height corresponds to median triplicate concentration for each metabolite, with the top 9 shown and the 44 other identified metabolites summed in grey. TMAO = trimethylamine N-oxide, HO-Ile = hydroxyisoleucine, GBT = glycine betaine, DCM = deep chlorophyll maximum. Lower row of boxplots shows the fraction of total particulate carbon (PC) in the known metabolites across the eddy dipole, with the upper whisker extent, median bar, and lower whisker extent corresponding to the values measured in three biological triplicates. Colors correspond to corrected sea level anomaly (Corr. SLA) in centimeters, with dark red indicating anticyclonic (positive SLA) and dark blue indicating cyclonic (negative SLA) eddy state. B) Type I linear regressions of particulate carbon against total nM carbon in known metabolites. Points have been colored according to the SLA scheme in A) and have been placed on independent axes. Outlier points have been removed from the surface at stations 6, 10, and 11 as well as a DCM sample from station 12. For the full regression, see Supplemental Figure XX.*

@@

We found that the overall concentration of known molecules was highest at the DCM of the cyclone, in line with our expectations and prior results. The most abundant quantified molecule overall was trimethylamine N-oxide (TMAO), with an average concentration of 1.38 nM and a distinct DCM maximum (Figure 2). This pattern of a subsurface maximum at the DCM was mirrored by a majority of the known molecules (37/53). Other abundant molecules decreased monotonically with depth, with highest values at the surface (Figure 2).

```{r PC_vs_totnM regression values, eval=FALSE}
no_outlier_frac_pc_in_metab_df %>%
  ggplot(aes(x=PC_um, y=total_nM)) + 
  geom_smooth(method="lm", color="grey40") +
  geom_point(aes(color=depth)) + 
  geom_smooth(aes(color=depth), method="lm") +
  theme_bw()

@@

As expected, bulk particulate carbon (PC) measurements showed a strong trend with summed metabolite concentrations ($\beta = 23.7$, Pearson's $r$ = .876, p-value < 0.001). This overall trend was largely driven by both higher values in the surface than at depth but when analyzed at each depth individually the general correspondence held. The DCM had the strongest trend ($\beta$ = 22.1, $r= .689$, p-value < 0.001) while Surface ($\beta$ = 16.0, $r = .375$, p-value = 0.04) and Deep ($\beta$ = 23.9, $r = 0.503$, p-value = 0.003) values were weaker but still compelling relationships. However, this association required the removal of three samples in the surface and one in the DCM with anomalously high metabolite loads. The full regression is reported in Supplemental Figure XX.

@@

Known metabolites contributed between 1.5% and 5% of the carbon to the total PC load with higher values at the DCM than the surface or 175m samples (Figure 2, Supplemental Figure XX). DCM samples contained approximately 25% (0.5 percentage points) more carbon in the known metabolite pool (One-way ANOVA with outliers removed plus Tukey post-hoc, p<0.001, n=30-33) but there was no clear trend with SLA.

@@

One sample, taken from the surface at Station 11 (MS11C215m_A), was highly distinct from all others (including its paired triplicates) with several metabolites measuring two or more orders of magnitude larger in peak area and concentration both before and after normalization to internal standards. Heuristic exploration of potentially related masses identified a total of 14 compounds highly abundant in this one sample alone. The only identified compounds with this trend were taurine and acetyltaurine, though three others with *m/z* 146.1176, 148.0968, and 206.1388 were discovered to be likely isomers of known compounds (acetylcholine/(3-carboxypropyl)trimethylammonium, hydroxyisoleucine, dexpanthenol, respectively). One feature (*m/z* = 184.0637, RT = 7.0 minutes) was identified as choline sulfate using a second run of standards, while an additional mass feature at 168.0689 and RT = 8.5 was putatively flagged as taurine betaine (same mass as hydroxyisoleucine). This mass was also one of the most highly enriched in the MS11C215m_A sample after taurine itself, with a peak area nearly 30x that of the next largest sample. No clear explanation emerged explaining why some compounds were hyperabundant in this surface sample, though one possible explanation was the *Trichodesmium* bloom occurring at the same time.

@@

stans <- read_csv("https://raw.githubusercontent.com/IngallsLabUW/Ingalls_Standards/master/Ingalls_Lab_Standards.csv") %>%
  filter(Column=="HILIC" & z>0) %>%
  select(cmpd_name=Compound_Name, formula=Empirical_Formula) %>%
  group_by(formula) %>%
  summarise(cmpd_name=paste(cmpd_name, collapse = ", "))
base_mz <- 85.089149355+1.007276
iter_mzs <- expand_grid(CH2=0:6, O=2:6) %>%
  mutate(mz=base_mz+CH2*14.01565+O*15.994915) %>%
  mutate(mz=round(mz, digits=5)) %>%
  mutate(cmd=paste0("mz%between%pmppm(", mz, ", 10)")) %>%
  mutate(formula=paste0("C", 5+CH2, "H", 11+CH2*2, "NO", O)) %>%
  left_join(stans) %>%
  mutate(label=ifelse(is.na(cmpd_name), formula, paste0(formula, " (", cmpd_name, ")"))) %>%
  mutate(casewhen_arg=paste0(cmd, "~'", label, "'"))
ms11_df <- c("C11H23NO6", "C11H23NO6", "C9H19NO5", "C8H17NO5", "C6H13NO6", "C8H17NO4") %>%
  data.frame(formula=., rtmed=c(8.5, 10, 8.5, 10.2, 5.2, 9.2)) %>%
  left_join(iter_mzs) %>%
  select(mzmed=mz, rtmed, formula, cmpd_name) %>%
  bind_rows(ms11_offenders) %>%
  arrange(mzmed) %>%
  select(mzmed, rtmed, formula, compound_name, feature) %>%
  mutate(formula=case_when(
    compound_name=="Taurine"~"C2H8NO3S",
    feature=="MS_POS_1050"~"C7H16NO2",
    compound_name=="Hydroxyisoleucine"~"C6H14NO3",
    compound_name=="N-Acetyltaurine"~"C4H10NO4S",
    feature=="MS_POS_1340"~"C5H14NO3S",
    feature=="MS_POS_1419"~"C8H18NO3",
    mzmed%between%pmppm(184.0637, 10)~"C5H14NO4S",
    mzmed%between%pmppm(198.0794, 10)~"C6H16NO4S",
    compound_name=="Dexpanthenol"~"C9H20NO4",
    TRUE~formula
  )) %>%
  mutate(compound_name=case_when(
    compound_name=="(3-Carboxypropyl)trimethylammonium?"~NA,
    compound_name=="Hydroxyisoleucine"~NA,
    compound_name=="Dexpanthenol"~NA,
    TRUE~compound_name
  )) %>%
  mutate(rtmed=round(rtmed, 1)) %>%
  mutate(compound_name=ifelse(is.na(compound_name), "", compound_name)) %>%
  mutate(feature=ifelse(is.na(feature), "", feature)) %>%
  dplyr::rename(`Median m/z`=mzmed, `Median RT`=rtmed, `Putative formula`=formula,
                `Compound name`=compound_name, `Feature ID`=feature)
knitr::kable(ms11_df)
```

We explored this *Trichodesmium* hypothesis further by analyzing additional samples taken from the bloom to see whether the compounds abundant in *Trichodesmium* were also abundant in samples taken directly from the bloom itself. Of the 14 masses identified in the environmental sample, 7 had good-quality peaks when viewed in the bloom samples as extracted ion chromatograms, though only 4 of these matched closely in retention time.

@@

Several molecule- or pathway-specific narratives emerged from the metabolomic data described above as well. The untargeted approach used here allowed us to describe and chatacterize signals from small molecules whose identity was unknown, a particularly promising approach in open ocean gyres where the largest fraction of unknowns is found [@Heal2021]. [Anitra: how many of your unknowns were also unknowns in Katherine's paper?] We found that this was especially true in anticyclonic eddies at all depths, with the diversity of the cyclone much better characterized by our list of authentic standards. Many of the trends detected below persisted even after normalizing within each sample, indicating that the shifts discussed have implications beyond simple scaling with biomass.

>>>>

High frequency, long-term observations at Station ALOHA in the North Pacific Subtropical Gyre (NPSG) over the past 25 years have revealed temporal and spatial variability in what had prevously been considered a homogenous environment [@Karl1999; @Karl2017; @Karl2021]. A major source of variability comes in the form of mesoscale eddies in which water is entrained into circular surface currents tens to hundreds of kilometers in diameter [@Karl2017; @McGillicuddy2016]. Prior research has established the importance of these mesoscale features in introducing nutrients into the euphotic zone and altering the community composition [@Sweeney2003; @Barone2019]. In particular, the depth and intensity of the deep chlorophyll maximum (DCM) at the top of the nitricline differs dramatically between cyclonic and anticyclonic eddies. @Barone2022 reported a shift in the microbial community at the DCM with eukaryotic phytoplankton enriched and prokaryotic phytoplankton depleted in cyclonic eddies, and @Hawco2021 and @Dugenne2023 showed that some of this difference was due to alterations in dissolved iron chemistry and nitrogen fixation.

@@

```{r figure stacked barplot MT eddy transect w PC and SLA}
avg_targ_concs <- targ_peaks %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MT") %>%
  filter(samp_type=="Smp") %>%
  group_by(compound_name) %>%
  mutate(mean_cmpd_nm=mean(nM)) %>%
  ungroup()
arrow_df <- data.frame(
  cyc_type=c("Anticyclone\ncenter", "Cyclone\ncenter"),
  x=c("6", "12"), xend=c("6", "12"),
  y=c(5, 4), yend=c(9.5, 6.5),
  color=c(scales::muted("red"), scales::muted("blue")),
  depth=factor("Deep", levels=c("Surface", "DCM", "Deep"), 
               labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)")),
  lab_x=c("5", "10")
)
barplot_targ_gp <- avg_targ_concs %>%
  group_by(compound_name, untripl) %>%
  summarise(nM=median(nM), mean_cmpd_nm=unique(mean_cmpd_nm)) %>%
  ungroup() %>%
  arrange(desc(mean_cmpd_nm)) %>%
  mutate(compound_name=ifelse(compound_name=="Trimethylamine N-oxide", "TMAO", compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Hydroxyisoleucine", "HO-Ile", compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Glycine betaine", "GBT", compound_name)) %>%
  mutate(compound_name=str_remove(compound_name, "^L-")) %>%
  mutate(compound_name=ifelse(
    compound_name%in%head(unique(compound_name), 9), 
    compound_name, "Other")) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  left_join(filled_file_metadata %>% distinct(untripl, depth, station)) %>%
  group_by(depth, station, compound_name) %>%
  summarise(nM=sum(nM), .groups = "keep") %>%
  summarise(mean_nM=mean(nM)) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)"))) %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_col(aes(x=station, y=mean_nM, fill=compound_name), width = 0.9) +
  scale_fill_manual(values = c(splividis(9), "grey50")) +
  scale_y_continuous(expand = c(0.005, 0.1)) +
  labs(x=NULL, y="Median nM\nmetabolite concentration", fill=NULL) +
  facet_grid(~depth, scales="free", space = "free") +
  guides(fill=guide_legend(nrow=2, byrow = TRUE)) +
  ggnewscale::new_scale_fill() +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend, color=color), data=arrow_df,
               arrow=arrow(ends = "first", length = unit(0.1, "npc"), type = "closed"),
               linewidth=0, linejoin = "bevel") +
  geom_segment(aes(x=x, y=y+0.5, xend=xend, yend=yend, color=color), 
               data=arrow_df, linewidth=2) +
  geom_text(aes(x=lab_x, y=yend+0.5, color=color, label=cyc_type), 
            data=arrow_df, vjust=0, hjust=0, lineheight=0.8) +
  scale_x_discrete(breaks=as.character(seq(4, 14, 2))) +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("fill"), guide = "none"
  ) +
  scale_color_identity() +
  theme_bw() +
  theme(legend.position = "top", axis.text.x = element_blank(), 
        legend.margin = margin(t = 10,0,0,0), 
        legend.box.margin = margin(-10,-10,-10,-10),
        plot.margin = unit(numeric(4), "cm"),
        strip.text = element_text(face = "bold"),
        plot.background = element_rect(fill="white"))

frac_pc_in_metab_df <- targ_peaks %>%
  filter(str_detect(filename, "180821_Smp")) %>%
  left_join(clean_stans %>% select(compound_name, formula, polarity)) %>%
  mutate(n_C=as.numeric(str_extract(formula, "(?<=C)\\d+"))) %>%
  mutate(nM_C=n_C*nM) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  filter(samp_type=="Smp") %>%
  group_by(filename) %>%
  summarize(total_nM=sum(nM_C), PC_um=unique(PC_um)) %>%
  mutate(frac_in_metab=total_nM/1000/PC_um) %>%
  left_join(filled_file_metadata, by = join_by(filename, PC_um)) %>%
  mutate(depth=str_extract(depth, "DCM|Deep|Surface")) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  mutate(station=factor(station, levels=c(as.character(4:14))))
pc_frac_metab_gp <- frac_pc_in_metab_df %>%
  ggplot(aes(x=station, y=frac_in_metab, color=sla_corr)) +
  geom_hline(yintercept = 0) +
  geom_boxplot(width = 0.9, linewidth=0.8) +
  scale_x_discrete(breaks=as.character(seq(4, 14, 2))) +
  scale_y_continuous(labels = scales::label_percent()) +
  facet_grid(~depth, scales="free") +
  scale_color_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("color", "fill")
  ) +
  guides(color=guide_colorbar(direction = "horizontal", title.position = "top", 
                              title.hjust = 0.5, barheight = 0.2, reverse = TRUE)) +
  labs(color="Corr. SLA (cm)", x="Station",
       y="Fraction of PC represented\nby known metabolites") +
  theme_bw() +
  theme(legend.position = c(0.84, 0.75), legend.background = element_rect(color="grey90"),
        strip.background = element_blank(), strip.text = element_blank(),
        plot.margin = margin(0,1,0,0), plot.background = element_rect(fill="white"))

@@

plot(targ_gp_w_sla_frac)

ggsave("targ_gp_w_sla_frac.png", plot = targ_gp_w_sla_frac, device = "png", 
       path = "figures", width = 6.875, height = 4.5, units = "in", dpi = 300, 
       bg = "white")
```

![](figures/targ_gp_w_sla_frac.png)

*Figure 2: Upper row of stacked barplots shows known compounds measured across the eddy dipole and separated by depth. Bar height corresponds to median triplicate concentration for each metabolite, with the top 9 shown and the 44 other identified metabolites summed in grey. TMAO = trimethylamine N-oxide, HO-Ile = hydroxyisoleucine, GBT = glycine betaine, DCM = deep chlorophyll maximum. Lower row of boxplots shows the fraction of total particulate carbon (PC) in the known metabolites across the eddy dipole, with the upper whisker extent, median bar, and lower whisker extent corresponding to the values measured in three biological triplicates. Colors correspond to corrected sea level anomaly (Corr. SLA) in centimeters, with dark red indicating anticyclonic (positive SLA) and dark blue indicating cyclonic (negative SLA) eddy state.*

@@

We found that the overall concentration of known molecules was highest at the DCM of the cyclone, in line with our expectations and prior results. The most abundant quantified molecule overall was trimethylamine N-oxide (TMAO), with an average concentration of 1.38 nM and a distinct DCM maximum (Figure 2). This pattern of a subsurface maximum at the DCM was mirrored by a majority of the known molecules (37/53). Most other abundant molecules decreased monotonically with depth, with highest values at the surface (Figure 2).

```{r PC_vs_totnM regression values, eval=FALSE}
no_outlier_frac_pc_in_metab_df <- frac_pc_in_metab_df %>%
  filter(!(station=="11" & tripl=="A" & depth=="Surface")) %>%
  filter(!(station=="6" & tripl=="B" & depth=="Surface")) %>%
  filter(!(station=="7" & tripl=="C" & depth=="Surface")) %>%
  filter(!(station=="11" & tripl=="B" & depth=="DCM"))

@@

The small-scale oceanographic variability introduced by mesoscale eddy features allowed us to also test the robustness of a model expecting a linear relationship between the total carbon contained within metabolites and bulk particulate carbon (PC) measurements. As expected, PC values showed a strong trend with summed metabolite concentrations ($\beta = 23.7\pm1.35$ (SE), Pearson's $r$ = .876, p-value < 0.001, Figure 2.5). This overall trend was largely driven by higher values in the surface than at depth, but when analyzed at each depth individually the general correspondence held. The DCM had the strongest trend ($\beta = 22.1\pm2.24$, $r= .689$, p-value < 0.001) while Surface ($\beta = 16.0\pm7.45$, $r = .375$, p-value = 0.04) and Deep ($\beta = 23.9\pm7.36$, $r = 0.503$, p-value = 0.003) values were weaker but still compelling relationships (Figure 2.5). However, this association required the removal of three samples in the surface and one in the DCM with anomalously high metabolite loads as well as two surface PC samples with anomalously high values from Stations 10 and 11 for which PC was interpolated by linear regression against beam attenuation. The full regression is reported in Supplemental Figure 3.

```{r Fig 2.5 PC vs tot nM plot}
pc_totalnm_plot <- no_outlier_frac_pc_in_metab_df %>%
  ggplot(aes(x=PC_um, y=total_nM)) +
  geom_point(aes(fill=depth), pch=21, size=3) +
  geom_smooth(method="lm", color="white", lwd=2.5, se=FALSE) +
  geom_smooth(method="lm", color="black", lwd=1.5) +
  geom_smooth(aes(group=depth), method="lm", color="white", lwd=2.5, se = FALSE) +
  geom_smooth(aes(color=depth), method="lm", lwd=1.5) +
  theme_bw() +
  theme(legend.position = "top") +
  labs(color="Sampling depth",
       fill="Sampling depth",
       x="Particulate carbon (uM)", 
       y="Total nM carbon in known metabolites") +
  scale_color_manual(breaks=c("Surface", "DCM", "Deep"), 
                     values = c("#F4BB23", "#028E34", "#0B505C"), aesthetics = c("fill", "color"))
pc_totalnm_plot_bydepth <- no_outlier_frac_pc_in_metab_df %>%
  ggplot() +
  aes(x=PC_um, y=total_nM) +
  geom_smooth(method="lm", color="black") +
  geom_point(aes(fill=sla_corr), pch=21, color="black") +
  facet_wrap(~depth, ncol=1, scales="free") +
  labs(x="Particulate\ncarbon (uM)", y="Total nM carbon in known metabolites") +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("fill"), name="Corr.\nSLA\n(cm)"
  ) +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))

comb_pcnm_plot <- cowplot::plot_grid(pc_totalnm_plot, pc_totalnm_plot_bydepth, 
                                     rel_widths = c(5, 3))

ggsave("comb_pcnm_plot.png", plot = comb_pcnm_plot, device = "png", 
       path = "figures", width = 6.875, height = 4.5, units = "in", dpi = 300, 
       bg = "white")
```

![](figures/comb_pcnm_plot.png)

*Figure 2.5: Type I linear regressions of particulate carbon against total nM carbon in known metabolites both as a whole and divided by depth. Points have been colored according to the SLA and have been placed on independent axes. Outlier metabolite points have been removed from the surface at stations 6, 10, and 11 as well as a DCM sample from station 12, while two outlier PC points have been interpolated from beam attenuation for surface values at stations 10 and 11. For the full regression, see Supplemental Figure 3.*

@@

Known metabolites contributed between 1.5% and 5% of the carbon to the total PC load with higher values at the DCM than the surface or 175m samples (Figure 2, Supplemental Figure 3). DCM samples contained approximately 25% (0.5 percentage points) more carbon in the known metabolite pool (One-way ANOVA with outliers removed plus Tukey post-hoc, p<0.001, n=30-33) but there was no clear trend with SLA.

@@

One sample, taken from the surface at Station 11 (MS11C215m_A), was highly distinct from all others (including its paired triplicates) with several metabolites measuring two or more orders of magnitude larger in peak area and concentration both before and after normalization to internal standards. Heuristic exploration of potentially related masses identified a total of 14 compounds highly abundant in this one sample alone. The only identified compounds with this trend were taurine and acetyltaurine, though three others with *m/z* 146.1176, 148.0968, and 206.1388 were discovered to be likely isomers of known compounds (acetylcholine/(3-carboxypropyl)trimethylammonium, hydroxyisoleucine, and dexpanthenol, respectively). One feature (*m/z* = 184.0637, RT = 7.0 minutes) was identified as choline sulfate using a second run of standards, while an additional mass feature at 168.0689 and RT = 8.5 was putatively flagged as taurine betaine. This mass was also one of the most highly enriched in the MS11C215m_A sample after taurine itself, with a peak area nearly 30x that of the next largest sample. No clear explanation emerged explaining why some compounds were hyperabundant in this surface sample, though one possible explanation was the *Trichodesmium* bloom occurring at the same time [@Dugenne2023].

@@

stans <- read_csv("https://raw.githubusercontent.com/IngallsLabUW/Ingalls_Standards/master/Ingalls_Lab_Standards.csv") %>%
  filter(Column=="HILIC" & z>0) %>%
  select(cmpd_name=Compound_Name, formula=Empirical_Formula) %>%
  group_by(formula) %>%
  summarise(cmpd_name=paste(cmpd_name, collapse = ", "))
base_mz <- 85.089149355+1.007276
iter_mzs <- expand_grid(CH2=0:6, O=2:6) %>%
  mutate(mz=base_mz+CH2*14.01565+O*15.994915) %>%
  mutate(mz=round(mz, digits=5)) %>%
  mutate(cmd=paste0("mz%between%pmppm(", mz, ", 10)")) %>%
  mutate(formula=paste0("C", 5+CH2, "H", 11+CH2*2, "NO", O)) %>%
  left_join(stans) %>%
  mutate(label=ifelse(is.na(cmpd_name), formula, paste0(formula, " (", cmpd_name, ")"))) %>%
  mutate(casewhen_arg=paste0(cmd, "~'", label, "'"))
ms11_df <- c("C11H23NO6", "C11H23NO6", "C9H19NO5", "C8H17NO5", "C6H13NO6", "C8H17NO4") %>%
  data.frame(formula=., rtmed=c(8.5, 10, 8.5, 10.2, 5.2, 9.2)) %>%
  left_join(iter_mzs) %>%
  select(mzmed=mz, rtmed, formula, cmpd_name) %>%
  bind_rows(ms11_offenders) %>%
  arrange(mzmed) %>%
  select(mzmed, rtmed, formula, compound_name, feature) %>%
  mutate(formula=case_when(
    compound_name=="Taurine"~"C2H8NO3S",
    feature=="MS_POS_1050"~"C7H16NO2",
    compound_name=="Hydroxyisoleucine"~"C6H14NO3",
    compound_name=="N-Acetyltaurine"~"C4H10NO4S",
    feature=="MS_POS_1340"~"C5H14NO3S",
    feature=="MS_POS_1419"~"C8H18NO3",
    mzmed%between%pmppm(184.0637, 10)~"C5H14NO4S",
    mzmed%between%pmppm(198.0794, 10)~"C6H16NO4S",
    compound_name=="Dexpanthenol"~"C9H20NO4",
    TRUE~formula
  )) %>%
  mutate(compound_name=case_when(
    compound_name=="(3-Carboxypropyl)trimethylammonium?"~NA,
    compound_name=="Hydroxyisoleucine"~NA,
    compound_name=="Dexpanthenol"~NA,
    formula=="C5H14NO4S"~"Choline sulfate",
    formula=="C5H14NO3S"~"Taurine betaine?",
    TRUE~compound_name
  )) %>%
  mutate(rtmed=round(rtmed, 1)) %>%
  mutate(compound_name=ifelse(is.na(compound_name), "", compound_name)) %>%
  mutate(feature=ifelse(is.na(feature), "", feature)) %>%
  dplyr::rename(`Median m/z`=mzmed, `Median RT`=rtmed, `Putative formula`=formula,
                `Compound name`=compound_name, `Feature ID`=feature)
knitr::kable(ms11_df)
```

| Median m/z| Median RT|Putative formula |Compound name    |Feature ID  |
|----------:|---------:|:----------------|:----------------|:-----------|
|   126.0221|      11.1|C2H8NO3S         |Taurine          |MS_POS_0677 |
|   146.1176|       8.5|C7H16NO2         |                 |MS_POS_1050 |
|   148.0968|       8.5|C6H14NO3         |                 |MS_POS_1080 |
|   168.0317|       4.9|C4H10NO4S        |N-Acetyltaurine  |MS_NEG_1123 |
|   168.0689|       8.5|C5H14NO3S        |Taurine betaine? |MS_POS_1340 |
|   176.1282|       8.1|C8H18NO3         |                 |MS_POS_1419 |
|   184.0637|       7.0|C5H14NO4S        |Choline sulfate  |            |
|   192.1230|       9.2|C8H17NO4         |                 |            |
|   196.0816|       5.2|C6H13NO6         |                 |            |
|   198.0794|       5.0|C6H16NO4S        |                 |            |
|   206.1388|       7.1|C9H20NO4         |                 |MS_POS_1683 |
|   208.1180|      10.2|C8H17NO5         |                 |            |
|   222.1336|       8.5|C9H19NO5         |                 |            |
|   266.1598|       8.5|C11H23NO6        |                 |            |
|   266.1598|      10.0|C11H23NO6        |                 |            |

*Table 1: Detected mass features with peak areas in sample MS11C215m_A much larger than the sample with the next-largest peak area. RT = retention time in minutes. Taurine betaine is followed by a ? to denote that this identification is putative. Feature ID is a disambiguating key produced by XCMS processing.*

@@

```{r}
raw_kheal <- read_csv("https://raw.githubusercontent.com/kheal/Gradients1_SemiTargeted3/master/Tables/Manuscript_tables/SuppTables/Full_EnviroCulture_PeakAreas.csv")
kheal_envirosamp <- raw_kheal %>%
  filter(Column=="HILIC") %>%
  mutate(polarity=ifelse(z>0, "pos", "neg")) %>%
  distinct(MassFeature_Column, compound_name_kath=Identification, Confidence, mz, rt, polarity) %>%
  mutate(rt=rt/60)
